Lagrangian Space Nonlinear $E$-mode clustering

### 2016-10-13 ms_disp1/ms_disp1.tex
figure captions
update figure
title, abstract, figures, box size

\addtolength{\hoffset}{-0.525cm}
\addtolength{\textwidth}{1.05cm}
\title{Reconstruction of Large Scale Structure by Nonlinear \tcr{Lagrangian} Displacement Fields}

@@

@@

\begin{abstract}
We investigate the ability of using nonlinear displacement field to reconstruct 
the primordial linear perturbations of the large scale structure. According to 
the linear Lagrangian perturbation theory, the first order primordial linear 
density field is given by the negative divergence of the displacement field. By 
running $N$-body simulations, we show that this \tcr{new} reconstruction algorithm recovers 
great amount of nonlinear information in the early universe. In reality, the 
displacement fields can be solved from only the final stage of the density 
fields. This demonstrates the potential to reconstruct baryonic acoustic oscillation (BAO) 
from current and future large scale structure surveys.

@@

\maketitle
\begin{figure}
 \centering
  \includegraphics[width=0.95\linewidth]{denx3.pdf}
  \caption{Visualization of the nonlinear density field $\delta_N$ (top),
  linear density field $\delta_L$ (middle) and the raw reconstructed density
  field $\delta_R$ (bottom). These projections have 9.375 Mpc$/h$ thickness
  and 300 Mpc$/h$ per side. The top panel shows the nonlinear displacement
  field $\bs\Psi$ by the deformed mesh, which traces the LSS of $\delta_N$.}
  \label{fig.projection}
\end{figure}
\section{Introduction}\label{sec.intro}
Our universe starts from primordial Gaussian perturbations at a very early stage, 
and from those fluctuations, the gravitational instability drives the formation of 
the large scale structure (LSS) distribution of matter
\citep{1970A&A.....5...84Z,1985ApJ...292..371D}.
These structures grow 
linearly until the perturbations are large enough so that the first order 
perturbation theories are unable to analytically describe the LSS distributions 
\citep{2016JCAP...01..043M}. 
As a result, the final nonlinear LSS distribution contains higher order 
statistics, and thus makes it more challenging to be interpreted into basic
cosmological parameters. One such example is that, the baryonic acoustic oscillation (BAO)
scale can be used as a ``standard ruler'' to constrain the cosmic expansion history
and thus probes the dark energy properties \citep{2005NewAR..49..360E}, but
nonlinear evolution smears the BAO features and lowers the measurement
accuracy \citep{2005ApJ...633..560E,2012MNRAS.419.2949N}.
There are various attempts to recover earlier stages
of LSS, in which statistics are closer to Gaussian 
\citep{1992MNRAS.254..315W,2013MNRAS.436..759H}.
Because Gaussian fields can be adequately described by two-point statistics,
ideally after some recovery algorithms, more information can be extracted,
more straightforwardly, by power spectra or two-point correlation functions
\citep{2005MNRAS.360L..82R,2012MNRAS.421..832Y}.

Standard BAO reconstruction algorithms smooth the 
nonlinear density field on linear scale ($\sim$10 Mpc$/h$) and reverse the large 
scale bulk flows by a negative Zel'dovich linear displacement
\citep{2007ApJ...664..675E,2009PhRvD..80l3501N,2009PhRvD..79f3523P}. Here we propose a 
\tcr{new reconstruction method} that uses the nonlinear displacement field to recover the 
primordial density field. In the linear Lagrangian perturbation theory, the 
negative divergence of the the displacement field
$\bs\Psi(\bs q)$ respect to Lagrangian coordinates $\bs q$
gives the linear density field \tcr{\citep{2010PhDT.........4J}}.
Of course, the full displacement field $\bs\Psi(\bs q)$
is non-observable, as it requires the initial distributions of matter, however there are 
many techniques to estimate the nonlinear displacement field from a final 
distribution of matter. For example, when a homogeneous initial matter distribution 
is assumed, there is a unique solution of curl-less displacement field to relate 
the initial and final distributions without shell-crossing. This solution can be 
solved by a metric transformation equation \citep{1995ApJS..100..269P,1998ApJS..115...19P}.
In 1-dimensional (1D) case, the exact solution of \cite{1995ApJS..100..269P}
simplifies to an ordering of matter elements by Eulerian coordinates.
Zhu {\it et al.} \cite{2016arXiv160907041Z} apply this algorithm to the result
of an 1D simulation \citep{2016JCAP...01..043M} and obtain an estimated
displacement field $\tilde\Psi(q)$, and find that \tcr{this new method} well recovers the linear
information and reconstructs BAO peak in correlation function. 
In 3D case, there are various techniques to obtain $\tilde{\bs \Psi}(\bs q)$.
However one needs to carefully consider effects of curl, shell-crossing in 3D case.

@@

\begin{figure}[t] \centering
  \includegraphics[width=1.0\linewidth]{fig2.pdf}
  \caption{Correlation functions $r(\delta_L,\delta_N)$ and $r(\delta_L,\delta_R)$
  (solid lines). Their
  damping of the power spectra $W^2$ (dashed lines),
  and their best fit BAO damping models (dotted lines).}
  \label{fig.corr}
\end{figure}

@@

\subsection{Simulation}\label{ss.sim}
We use the open source cosmological simulation code {\tt CUBE} %\footnote{github link of cafcube}
\citep{cafcube}.
Cosmological parameters are set as
$\Omega_m=0.27$, $\Omega_\Lambda=0.73$, $h_0=0.68$, $n_s=0.96$ and $\sigma_8=0.83$.
Initial conditions are generated at redshift $z=50$ 
using Zel'dovich approximation, and using a CLASS transfer function.
$N_p=512^3$ $N$-body particles are evolved via 
their mutual gravitational interactions to $z=0$, in a periodic box with $L=300$ 
Mpc$/h$ per side. The code is set to use standard a particle-mesh (PM) algorithm 
\cite{1988csup.book.....H} on a two-level mesh grids
(details see \cite{2013MNRAS.436..540H}) and cloud-in-cell
(CIC) is used in particle interpolations in force 
calculation and obtaining the density field $\rho({\bs x})$ in Eulerian coordinates 
${\bs x}$ at late stages. We use density contrast $\delta\equiv\rho/\lb\rho\rb-1$ 
to describe the density fluctuations. The primordial linear density field $
\delta_L$ is given by the initial stage and scaled to $z=0$ by the linear growth 
factor. In the top and middle panels of Fig.1 we show projections of the nonlinear density field
$\delta_N$ given by the simulation and the linear density field $\delta_L$.
$\delta_N$ is obtained by the particle distribution at redshift $z=0$, and
the particles are interpolated by using the cloud-in-cell (CIC) algorithm.
Because $\delta_N$ is highly nonlinear and follows an approximate
log-normal distribution, we plot $\log_{10}(\delta_N+1)$ instead.
The nonlinear evolution of $\delta_N$ makes it very different from $\delta_L$
in apprearence.

@@

\begin{figure}[t] \centering
  \includegraphics[width=1.0\linewidth]{recopower.pdf}
  \caption{Power spectra of the nonlinear density field and its
  filtered density field.}
  \label{fig.recopower}
\end{figure}

\begin{figure}[t] \centering
  \includegraphics[width=1.0\linewidth]{pdfs.pdf}
  \caption{Power spectra of the reconstructed density field and its
  filtered density field.}
  \label{fig.pdfs}
\end{figure}

@@

Fig.\ref{fig.corr} shows the cross correlation functions,
damping factors $W^2(k)$ for the optimal filtered 
nonlinear and reconstructed density fields.
We can see that the cross correlation $r_{LR}$ is recovered to
nearly an order of magnitude smaller scales compared to $r_{LN}$.
We also fit the Gaussian BAO damping model $
{\mathcal D}(k)=\exp(-k^2\Sigma^2/2)$ and give $\Sigma=1.8$ Mpc$/h$ and $\Sigma=10$ 
Mpc$/h$ for reconstructed and nonlinear fields.
We repeat the analysis with various box sizes,
and the results are consistent.

@@

In Fig.\ref{fig.pdfs} we use the probability distribution
as functions of $(\delta_L,\delta_N)$ and $(\delta_L,\delta_R)$
to show the point-point correlation between these two pairs of density
fields. In the top panel, $\delta_N$ shows an approximate log-normal
distribution, and shows no correlation with $\delta_L$. Although
$\delta_L$ and $\delta_N$ have correlations on linear scales (Fig.\ref{fig.corr}),
they have no correlation in real space, because initial density fluctuations
in Lagrangian coordinates are evolved/transformed to Eulerian coordinates.
As the reconstruction is done in Lagrangian space, it recovers
certain amount of correlation as shown in the bottom panel
of Fig.\ref{fig.pdfs}. One can also see that $\delta_R$ follows
a much closer Gaussian distribution.
\tcr{We notice that the less-dense regions of $\delta_L$ are more
correlated with $\delta_R$, whereas denser regions of $\delta_L$ are
reconstructed worse, and $\delta_R$ shows more dispersion.}

\section{Discussion and conclusion}\label{sec.discussion}
We extract the actual displacement field of matter elements in cosmological $N$-body
simulations, and use this displacement field in Lagrangian coordinates to reconstruct
the primordial linear perturbations. The result shows a prominent improvement from
$r_{LN}$ to $r_{LR}$ in Fig.\ref{fig.corr} -- recovering the lost linear information on
nearly an order of magnitude smaller scales.
This is achieved by implementing differential movement information
of matter elements on Lagrangian coordinates, rather than on
Eulerian coordinates. This result illustrates the feasibility
of using estimated displacement field $\tilde{\bs \Psi}(\bs q)$ to reconstruct primordial linear
density field. A straightforward example of a estimation of $\tilde{\bs \Psi}(\bs q)$
is given by \cite{1995ApJS..100..269P,1998ApJS..115...19P}.
\tcr{More complicated reconstruction algorithms, e.g., least action
algorithms \cite{1989ApJ...344L..53P,2000ApJ...544...21G} can give
the exact $\delta_L$}, where
shell-crossing effects are also considered. In reality, one needs
to consider all aspects including vorticity, shell-crossing, bias, noise
and data complexities. The impact of these factors can be quantitatively
compared with the impact of different estimation methods, and with
the exact solution by $N$-body simulations.

\tcr{A proper name of this method (Lagrangian displacement)?}

\tcr{The advantage of using displacement field in reconstruction is
its insensitive response from highly nonlinearities --
densest regions form because matter elements, after shell crossing,
stop experiencing the linear extrapolated of Zel'dovich approximation.
i.e. the displacement fields are dominated by early stage linear processes,
which is the Lagrangian-Eulerian coordinate transform,
while late stage shell-crossing, nonlinear and baryonic processes
only fine-tunes the final position $\bs x$. In contrast, traditional
treatments in reconstruction deals directly on density fields which sensitively
relies on nonlinear processes --  density values can vary by orders
of magnitude due to nonlinear/baryonic physics and many sources of errors.}

\tcr{Advantages of this method (simpler?) over least action?}

\section*{Acknowledgements}
We thank Hong-Ming Zhu for helpful discussions.
HRY acknowledges General Financial Grant No.2015M570884 and Special Financial Grant No.2016T90009 from the China Postdoctoal Science Foundation.

>>>>

\addtolength{\hoffset}{-0.525cm}
\addtolength{\textwidth}{1.05cm}
\title{Lagrangian Space Nonlinear $E$-mode clustering}

@@

\author{Hong-Ming Zhu}%email{hmzhu@bao.ac.cn}
\affiliation{Key Laboratory for Computational Astrophysics, National Astronomical Observatories, Chinese Academy of Sciences, 20A Datun Road, Beijing 100012, China}
\affiliation{University of Chinese Academy of Sciences, Beijing 100049, China}

@@

\begin{abstract}
We study the nonlinear $E$-mode clustering in Lagrangian space
by using large scale structure (LSS) $N$-body simulations
and use the displacement field information in Lagrangian space
to estimate the primordial linear density field.
We find that, compared to Eulerian nonlinear density fields,
the negative divergence of $E$-mode displacement fields
in Lagrangian space improves the cross-correlation with
initial density field by factor of 6 $\sim$ 7,
containing 2 orders of magnitude more primordial information. 
In reality, the displacement field is dominated by linear evolution
and insensitive of nonlinear shell-crossing, vorticity or
baryonic physics, and thus can be stably solved from only
the final matter distributions.
This illustrates ability of potential reconstruction algorithms,
to reconstruct baryonic acoustic oscillation (BAO) 
from current and future large scale structure surveys.

@@

\maketitle
\begin{figure}
 \centering
  \includegraphics[width=0.95\linewidth]{proj.pdf}
  \caption{Visualization of the nonlinear density field $\delta_N$ (top),
  linear density field $\delta_L$ (middle) and the raw reconstructed density
  field $\delta_R$ (bottom). These projections have 9.375 Mpc$/h$ thickness
  and 400 Mpc$/h$ per side. The top panel shows the nonlinear displacement
  field $\bs\Psi$ by the deformed mesh, which traces the LSS of $\delta_N$.}
  \label{fig.projection}
\end{figure}
\section{Introduction}\label{sec.intro}
Our universe starts from primordial Gaussian perturbations at a very early stage, 
and from those fluctuations, the gravitational instability drives the formation of 
the large scale structure (LSS) distribution of matter
\citep{1970A&A.....5...84Z,1985ApJ...292..371D}.
These structures grow 
linearly until the perturbations are large enough so that the first order 
perturbation theories are unable to analytically describe the LSS distributions 
\citep{2016JCAP...01..043M}. 
As a result, the final nonlinear LSS distribution contains higher order 
statistics, and thus makes it more challenging to be interpreted into basic
cosmological parameters. One such example is that, the baryonic acoustic oscillation (BAO)
scale can be used as a ``standard ruler'' to constrain the cosmic expansion history
and thus probes the dark energy properties \citep{2005NewAR..49..360E}, but
nonlinear evolution smears the BAO features and lowers the measurement
accuracy \citep{2005ApJ...633..560E,2012MNRAS.419.2949N}.
There are various attempts to recover earlier stages
of LSS, in which statistics are closer to Gaussian 
\citep{1992MNRAS.254..315W,2013MNRAS.436..759H}.
Because Gaussian fields can be adequately described by two-point statistics,
ideally after some recovery algorithms, more information can be extracted,
more straightforwardly, by power spectra or two-point correlation functions
\citep{2005MNRAS.360L..82R,2012MNRAS.421..832Y}.

Standard BAO reconstruction algorithms smooth the 
nonlinear density field on linear scale ($\sim$10 Mpc$/h$) and reverse the large 
scale bulk flows by a negative Zel'dovich linear displacement
\citep{2007ApJ...664..675E,2009PhRvD..80l3501N,2009PhRvD..79f3523P}. Here we propose a 
new reconstruction method that uses the nonlinear displacement field to recover the 
primordial density field. In the linear Lagrangian perturbation theory, the 
negative divergence of the the displacement field
$\bs\Psi(\bs q)$ respect to Lagrangian coordinates $\bs q$
gives the linear density field \citep{2010PhDT.........4J}.
Of course, the full displacement field $\bs\Psi(\bs q)$
is non-observable, as it requires the initial distributions of matter, however there are 
many techniques to estimate the nonlinear displacement field from a final 
distribution of matter. For example, when a homogeneous initial matter distribution 
is assumed, there is a unique solution of curl-less displacement field to relate 
the initial and final distributions without shell-crossing. This solution can be 
solved by a metric transformation equation \citep{1995ApJS..100..269P,1998ApJS..115...19P}.
In 1-dimensional (1D) case, the exact solution of \cite{1995ApJS..100..269P}
simplifies to an ordering of matter elements by Eulerian coordinates.
Zhu {\it et al.} \cite{2016arXiv160907041Z} apply this algorithm to the result
of an 1D simulation \citep{2016JCAP...01..043M} and obtain an estimated
displacement field $\tilde\Psi(q)$, and find that this new method well recovers the linear
information and reconstructs BAO peak in correlation function. 
In 3D case, there are various techniques to obtain $\tilde{\bs \Psi}(\bs q)$.
However one needs to carefully consider effects of curl, shell-crossing in 3D case.

@@

\begin{figure}[t] \centering
  \includegraphics[width=1.0\linewidth]{filter.pdf}
  \caption{Correlation functions $r(\delta_L,\delta_N)$ and $r(\delta_L,\delta_R)$
  (solid lines) and their scaled BAO damping models (dotted lines).}
  \label{fig.corr}
\end{figure}

@@

\subsection{Simulation}\label{ss.sim}
We use the open source cosmological simulation code {\tt CUBE} %\footnote{github link of cafcube}
\citep{cafcube}.
Cosmological parameters are set as
$\Omega_m=0.27$, $\Omega_\Lambda=0.73$, $h_0=0.68$, $n_s=0.96$ and $\sigma_8=0.83$.
Initial conditions are generated at redshift $z=50$ 
using Zel'dovich approximation, and using a CLASS transfer function.
$N_p=512^3$ $N$-body particles are evolved via 
their mutual gravitational interactions to $z=0$, in a periodic box with $L=400$ 
Mpc$/h$ per side. The code is set to use standard a particle-mesh (PM) algorithm 
\cite{1988csup.book.....H} on a two-level mesh grids
(details see \cite{2013MNRAS.436..540H}) and cloud-in-cell
(CIC) is used in particle interpolations in force 
calculation and obtaining the density field $\rho({\bs x})$ in Eulerian coordinates 
${\bs x}$ at late stages. We use density contrast $\delta\equiv\rho/\lb\rho\rb-1$ 
to describe the density fluctuations. The primordial linear density field $
\delta_L$ is given by the initial stage and scaled to $z=0$ by the linear growth 
factor. In the top and middle panels of Fig.1 we show projections of the nonlinear density field
$\delta_N$ given by the simulation and the linear density field $\delta_L$.
$\delta_N$ is obtained by the particle distribution at redshift $z=0$, and
the particles are interpolated by using the cloud-in-cell (CIC) algorithm.
Because $\delta_N$ is highly nonlinear and follows an approximate
log-normal distribution, we plot $\log_{10}(\delta_N+1)$ instead.
The nonlinear evolution of $\delta_N$ makes it very different from $\delta_L$
in apprearence.

@@

\begin{figure}[t] \centering
  \includegraphics[width=1.0\linewidth]{power.pdf}
  \caption{Power spectra decomposition of density fields and their
  filtered density field.}
  \label{fig.recopower}
\end{figure}

\begin{figure}[t] \centering
  \includegraphics[width=1.0\linewidth]{pdfs.pdf}
  \caption{2D-PDF (preliminary)}
  \label{fig.pdfs}
\end{figure}

@@

Fig.\ref{fig.corr} shows the cross correlation functions,
damping factors $W^2(k)$ for the optimal filtered 
nonlinear and reconstructed density fields.
We can see that the cross correlation $r_{LR}$ is recovered to
nearly an order of magnitude smaller scales compared to $r_{LN}$.
We also fit the Gaussian BAO damping model $
{\mathcal D}(k)=\exp(-k^2\Sigma^2/2)$ and give $\Sigma=1.8$ Mpc$/h$ and $\Sigma=12$ 
Mpc$/h$ for reconstructed and nonlinear fields.
We repeat the analysis with various box sizes,
and the results are consistent.

@@

In Fig.\ref{fig.pdfs} we use the probability distribution
as functions of $(\delta_L,\delta_N)$ and $(\delta_L,\delta_R)$
to show the point-point correlation between these two pairs of density
fields. In the top panel, $\delta_N$ shows an approximate log-normal
distribution, and shows no correlation with $\delta_L$. Although
$\delta_L$ and $\delta_N$ have correlations on linear scales (Fig.\ref{fig.corr}),
they have no correlation in real space, because initial density fluctuations
in Lagrangian coordinates are evolved/transformed to Eulerian coordinates.
As the reconstruction is done in Lagrangian space, it recovers
certain amount of correlation as shown in the bottom panel
of Fig.\ref{fig.pdfs}. One can also see that $\delta_R$ follows
a much closer Gaussian distribution.
We notice that the less-dense regions of $\delta_L$ are more
correlated with $\delta_R$, whereas denser regions of $\delta_L$ are
reconstructed worse, and $\delta_R$ shows more dispersion.
This is caused by nonlinear effects and it damps out in
higher redshifts.

\section{Discussion and conclusion}\label{sec.discussion}
We extract the actual displacement field of matter elements in cosmological $N$-body
simulations, and use this displacement field in Lagrangian coordinates to reconstruct
the primordial linear perturbations. The result shows a prominent improvement from
$r_{LN}$ to $r_{LR}$ in Fig.\ref{fig.corr} -- recovering the lost linear information on
nearly an order of magnitude smaller scales.
This is achieved by implementing differential movement information
of matter elements on Lagrangian coordinates, rather than on
Eulerian coordinates. This result illustrates the feasibility
of using estimated displacement field $\tilde{\bs \Psi}(\bs q)$ to reconstruct primordial linear
density field. A straightforward example of a estimation of $\tilde{\bs \Psi}(\bs q)$
is given by \cite{1995ApJS..100..269P,1998ApJS..115...19P}.
%\tcr{More complicated reconstruction algorithms, e.g., least action
%algorithms \cite{1989ApJ...344L..53P,2000ApJ...544...21G} can give
%the exact $\delta_L$}, where
%shell-crossing effects are also considered.
In reality, one needs
to consider all aspects including vorticity, shell-crossing, bias, noise
and data complexities. The impact of these factors can be quantitatively
compared with the impact of different estimation methods, and with
the exact solution by $N$-body simulations.

The advantage of using displacement field in reconstruction is
its insensitive response from highly nonlinearities --
densest regions form because matter elements, after shell crossing,
stop experiencing the linear extrapolated of Zel'dovich approximation.
i.e. the displacement fields are dominated by early stage linear processes,
which is the Lagrangian-Eulerian coordinate transform,
while late stage shell-crossing, nonlinear and baryonic processes
only fine-tunes the final position $\bs x$. In contrast, traditional
treatments in reconstruction deals directly on density fields which sensitively
relies on nonlinear processes --  density values can vary by orders
of magnitude due to nonlinear/baryonic physics and many sources of errors.

\section*{Acknowledgements}
We thank Pengjie Zhang for helpful discussions which triggered
the inspiration of the work.
HRY acknowledges General Financial Grant No.2015M570884 and Special Financial Grant No.2016T90009 from the China Postdoctoal Science Foundation.
[ZHM's acknowledgements.]
The Dunlap Institute is funded through an endowment established by the David Dunlap family and the University of Toronto.
Research at the Perimeter Institute is supported by the Government of Canada
through Industry Canada and by the Province of Ontario through the Ministry of
Research $\&$ Innovation.
